---
title: "Vignette for `dampack` package"
author: "Fernando Alarid Escudero"
date: "`r Sys.Date()`"
output: 
  - rmarkdown::html_vignette
  # - pdf_document
  #BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: /Users/FAE/Dropbox/BibTeX/Statistics.bib
link-citations: true
#csl: sage-vancouver.csl #elsevier-vancouver.csl
---

## Estimation of Dirichlet parameters using MoM
This function returns the $\alpha$ parameteres of a dirichlet distribution following the method of moments (MoM) proposed by @Fielitz1975 and @Narayanan1992.

If $\mu$ is a vector of means and $\sigma$ is a vector of standard deviations of the random variable, then the second moment $X_2$ is defined by $\sigma^2 + \mu^2$. Using the mean and the second moment, the $J$ alpha parameters are computed as follows
$$ \alpha_i = \frac{(\mu_1-X_{2_{1}})\mu_i}{X_{2_{1}}-\mu_1^2} $$
for $i = 1, \ldots, J-1$, and

$$ \alpha_J = \frac{(\mu_1-X_{2_{1}})(1-\sum_{i=1}^{J-1}{\mu_i})}{X_{2_{1}}-\mu_1^2} $$

## Estimation of Log-normal parameters uisng MoM
This function returns the location, $\mu$, and scale, $\sigma$, parameteres of a log-normal distribution from the mean and variance of a random variable following the method of moments (MoM).

Given the non-logarithmized mean and variance $m$ and $v$ of the random variable, respectively, the location, $\mu$, and scale, $\sigma$, of a log-normal distribution are given by the following equations

$$ \mu = \ln{\left(\frac{m}{\sqrt{\left(1 + \frac{v}{m^2} \right)}}\right)} $$
and

$$ \sigma = \sqrt{\ln{\left( 1 + \frac{v}{m^2}\right)}} $$

## EVPPI using linear regression metamodeling
The expected value of partial pefect information (EVPPI) is the expected value of perfect information from a subset of parameters of interest, $\theta_I$ of a cost-effectiveness analysis (CEA) of $D$ different strategies with parameters $\theta = \{ \theta_I, \theta_C\}$, where $\theta_C$ is the set of complimenatry parameters of the CEA. The function `evppi_lrmm` computes the EVPPI of $\theta_I$ from a matrix of net monetary benefits $B$ of the CEA. Each column of $B$ corresponds to the net benefit $B_d$ of strategy $d$. The function `evppi_lrmm` computes the EVPPI using a linear regression metamodel [@Strong2014b; @Jalal2017] approach following these steps:

1. Determine the optimal strategy $d^*$ from the expected net benefits $\bar{B}$
$$ d^* = \operatorname{arg\,max}_{d} \left\{\bar{B}\right\} $$
1. Compute the opportunity loss for each $d$ strategy, $L_d$
$$ L_d = B_d - B_{d^*} $$
1. Estimate a linear metamodel for the opportunity loss of each $d$ strategy, $L_d$, by regressing them on the spline basis functions of $\theta_I$, $f(\theta_I)$
$$ L_d = \beta_0 + f(\theta_I) + \epsilon,$$
where $\epsilon$ is the residual term that captures the complementary parameters $\theta_C$ and the difference between the original simulation model and the metamodel.
1. Compute the EVPPI of $\theta_I$ using the estimated losses for each $d$ strategy, $\hat{L}_d$ from the linear regression metamodel and applying the following equation:
$$ \text{EVPPI}_{\theta_I} = \frac{1}{K}\sum_{i=1}^{K}\max_d\left(\hat{L}_d \right)$$

The spline model in step 3 is fitted using the `mgcv` package.

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))

# References